The topos-theoretic multiverse: \\ a modal approach for computation

### 2023-02-12 paper-modal-multiverse.tex
to do
Switch to llncs style for submission to CiE 2023

\documentclass[oneside,reqno]{amsart}

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,stmaryrd,amssymb}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{tikz}
\usepackage{proof}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{color}

\usepackage[natbib=true,style=numeric,maxnames=10]{biblatex}
\usepackage[babel]{csquotes}
\bibliography{paper-modal-multiverse.bib}

\title{The topos-theoretic multiverse: a modal approach}
\author{Ingo Blechschmidt}
\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}
\email{ingo.blechschmidt@math.uni-augsburg.de}

\theoremstyle{definition}
\newtheorem{defn}{Definition}[section]
\newtheorem{ex}[defn]{Example}

\theoremstyle{plain}
\newtheorem{prop}[defn]{Proposition}
\newtheorem{cor}[defn]{Corollary}
\newtheorem{lemma}[defn]{Lemma}
\newtheorem{thm}[defn]{Theorem}
\newtheorem{scholium}[defn]{Scholium}

\theoremstyle{remark}
\newtheorem{rem}[defn]{Remark}
\newtheorem{question}[defn]{Question}
\newtheorem{speculation}[defn]{Speculation}
\newtheorem{caveat}[defn]{Caveat}
\newtheorem{conjecture}[defn]{Conjecture}

@@

@@

\maketitle
\thispagestyle{empty}

@@

\textbf{Acknowledgments.} Alexander Oldenziel, \ldots

@@

\begin{defn}A \emph{Grothendieck topos} is a category equivalent to
the category of sheaves on a small site. A \emph{Grothendieck topos over a
given (Grothendieck or elementary) topos}~$\B$ is ``a Grothendieck topos from
the point of view of~$\B$''; using a sufficiently expressive form of the
internal language of~$\B$, which allows us to make direct sense of the notion in
scare quotes, this amounts to a bounded geometric morphism~$\E \to
\B$, and this notion can be taken as the definition.
\end{defn}

@@

\begin{rem}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
fibrations/indexed categories over~$\B$ (XXX: are these the same?) validating an appropriate form of the
axioms of elementary toposes, should also be
taken as part of the topos-theoretic multiverse. The first foray into the
modal topos-theoretic multiverse outlined in this note
sticks to Grothendieck toposes for ease of
formalizability (``for every small site'' can be expressed also in the more
standard flavors of the internal topos language, while ``for every elementary
topos'' requires more elaborate versions); because the restricted multiverse
is already sufficiently rich for the intended applications; and because we have
a generalization to the predicative setting~\cite{crosilla:xxx} with arithmetic
universes in mind. Predicatively, not even the category of sets might be an
elementary topos, but the category of sets and categories of sheaves are still
arithmetic universes.
\end{rem}

@@

\begin{defn}A topos~$\E$ is \emph{positive} if and only if the geometric
morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
isomorphisms).\end{defn}

\begin{ex}The topos of sheaves over a topological space~$X$ is positive if and
only if~$X$ has a point; more generally, the topos of sheaves over a locale~$X$
is positive if and only if every open covering of the top element of the frame
of~$X$ is inhabited. As such, positivity is a more informative
version of the constructively weaker property of being nontrivial (the property that the top open
and the bottom open do not coincide).\end{ex}

\begin{ex}The spectrum of a ring~$A$, that is the classifying topos of prime
filters of~$A$ (or equivalently the topos of sheaves over the classifying
locale of prime filters of~$A$) is positive iff~$1 \neq 0$ in~$A$.\end{ex}

\begin{ex}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ to be positive is that xxx.\end{ex}

@@

\begin{defn}A topos~$\E$ is \emph{ouvert} if and only if the geometric
morphism~$f : \E \to \Set$ is open, that is~$f^*$ preserves the interpretation of
first-order formulas.\end{defn}

\begin{ex}The topos of sheaves over a topological space is always ouvert. More
generally, the topos of sheaves over a locale~$X$ is ouvert iff there exists a
\emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{ex}

\begin{ex}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{ex}

\begin{ex}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ being ouvert is that the indexing sets of all disjunctions
appearing on the right hand side of turnstiles, in a normal form presentation
of~$\TT$, are inhabited~\cite{xxx}.\end{ex}

@@

\begin{prop}\label{prop:gen-good}Let~$R$ be a relation on a set~$X$. The
generic infinite sequence over~$X$ is good if and only if~$R$ is almost-full in
the inductive sense.\end{prop}

@@

\begin{cor}\label{cor:modal-good}Let~$R$ be a relation on a set~$X$. Then~$R$
is almost-full in the inductive sense if and only if everywhere, every infinite
sequence is good.\end{cor}

@@

\begin{prop}Let~$({<})$ be a transitive relation on a set~$X$. The generic
infinite descending chain over~$X$ validates~$\bot$ (that is, the classifying
topos of such chains is trivial) if and only if the relation is well-founded in
the inductive sense.\end{prop}

@@

\begin{cor}A transitive relation is well-founded in the inductive sense
if and only if everywhere, it is not the case that there exists an infinite
descending chain.\end{cor}

@@

\begin{prop}\label{prop:af-wf}Let~$({\leq})$ be a transitive almost-full relation.
Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
is well-founded.\end{prop}

@@

\begin{rem}For the proof of Proposition~\ref{prop:af-wf}, it is not relevant
that pullback along geometric morphisms typically fails to preserve the
negation occuring in the definition of~$({<})$, basically because we still
have~$f^*(\llbracket x > y \rrbracket) \wedge f^*(\llbracket x \leq y
\rrbracket) \Rightarrow f^*(\llbracket\bot\rrbracket) = \bot$.\end{rem}

\begin{thm}[Dickson's lemma] If~$X$ and~$Y$ are almost-full,
so is~$X \times Y$.\end{thm}

@@

    \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
    among all values of~$\beta \circ i$. Hence~$\gamma$ is good in view of
    \begin{align*}
      \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
      \beta(i(k_0)) &\leq \beta(i(k_0+1)). \qedhere
    \end{align*}
  \end{enumerate}
\end{proof}

@@

@@

>>>>

\documentclass[envcountsect,envcountsame,runningheads]{llncs}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath,mathtools,stmaryrd,amssymb}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{tikz}
\usepackage{proof}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{color}

\title{The topos-theoretic multiverse: \\ a modal approach for computation}

@@

\maketitle

@@

@@

\textbf{Acknowledgments.} Alexander Oldenziel, \ldots
% xxx: Iosif

@@

\begin{definition}A \emph{Grothendieck topos} is a category equivalent to
the category of sheaves on a small site. A \emph{Grothendieck topos over a
given (Grothendieck or elementary) topos}~$\B$ is ``a Grothendieck topos from
the point of view of~$\B$''; using a sufficiently expressive form of the
internal language of~$\B$, which allows us to make direct sense of the notion in
scare quotes, this amounts to a bounded geometric morphism~$\E \to
\B$, and this notion can be taken as the definition.
\end{definition}

@@

\begin{remark}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
fibrations/indexed categories over~$\B$ (XXX: are these the same?) validating an appropriate form of the
axioms of elementary toposes, should also be
taken as part of the topos-theoretic multiverse. The first foray into the
modal topos-theoretic multiverse outlined in this note
sticks to Grothendieck toposes for ease of
formalizability (``for every small site'' can be expressed also in the more
standard flavors of the internal topos language, while ``for every elementary
topos'' requires more elaborate versions); because the restricted multiverse
is already sufficiently rich for the intended applications; and because we have
a generalization to the predicative setting~\cite{crosilla:xxx} with arithmetic
universes in mind. Predicatively, not even the category of sets might be an
elementary topos, but the category of sets and categories of sheaves are still
arithmetic universes.
\end{remark}

@@

\begin{definition}A topos~$\E$ is \emph{positive} if and only if the geometric
morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
isomorphisms).\end{definition}

\begin{example}The topos of sheaves over a topological space~$X$ is positive if and
only if~$X$ has a point; more generally, the topos of sheaves over a locale~$X$
is positive if and only if every open covering of the top element of the frame
of~$X$ is inhabited. As such, positivity is a more informative
version of the constructively weaker property of being nontrivial (the property that the top open
and the bottom open do not coincide).\end{example}

\begin{example}The spectrum of a ring~$A$, that is the classifying topos of prime
filters of~$A$ (or equivalently the topos of sheaves over the classifying
locale of prime filters of~$A$) is positive iff~$1 \neq 0$ in~$A$.\end{example}

\begin{example}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ to be positive is that xxx.\end{example}

@@

\begin{definition}A topos~$\E$ is \emph{ouvert} if and only if the geometric
morphism~$f : \E \to \Set$ is open, that is~$f^*$ preserves the interpretation of
first-order formulas.\end{definition}

\begin{example}The topos of sheaves over a topological space is always ouvert. More
generally, the topos of sheaves over a locale~$X$ is ouvert iff there exists a
\emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{example}

\begin{example}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{example}

\begin{example}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ being ouvert is that the indexing sets of all disjunctions
appearing on the right hand side of turnstiles, in a normal form presentation
of~$\TT$, are inhabited~\cite{xxx}.\end{example}

@@

\begin{proposition}\label{prop:gen-good}Let~$R$ be a relation on a set~$X$. The
generic infinite sequence over~$X$ is good if and only if~$R$ is almost-full in
the inductive sense.\end{proposition}

@@

\begin{corollary}\label{cor:modal-good}Let~$R$ be a relation on a set~$X$. Then~$R$
is almost-full in the inductive sense if and only if everywhere, every infinite
sequence is good.\end{corollary}

@@

\begin{proposition}Let~$({<})$ be a transitive relation on a set~$X$. The generic
infinite descending chain over~$X$ validates~$\bot$ (that is, the classifying
topos of such chains is trivial) if and only if the relation is well-founded in
the inductive sense.\end{proposition}

@@

\begin{corollary}A transitive relation is well-founded in the inductive sense
if and only if everywhere, it is not the case that there exists an infinite
descending chain.\end{corollary}

@@

\begin{proposition}\label{prop:af-wf}Let~$({\leq})$ be a transitive almost-full relation.
Then~$({<})$, where~$x < y \equiv (x \leq y \wedge \neg(y \leq x))$,
is well-founded.\end{proposition}

@@

\begin{remark}For the proof of Proposition~\ref{prop:af-wf}, it is not relevant
that pullback along geometric morphisms typically fails to preserve the
negation occuring in the definition of~$({<})$, basically because we still
have~$f^*(\llbracket x > y \rrbracket) \wedge f^*(\llbracket x \leq y
\rrbracket) \Rightarrow f^*(\llbracket\bot\rrbracket) = \bot$.\end{remark}

\begin{theorem}[Dickson's lemma] If~$X$ and~$Y$ are almost-full,
so is~$X \times Y$.\end{theorem}

@@

    \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
    among all values of~$\beta \circ i$. Hence~$\gamma$ is good in view of
    \begin{align*}
      \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
      \beta(i(k_0)) &\leq \beta(i(k_0+1)).
    \end{align*}
  \end{enumerate}
\end{proof}

@@

\bibliographystyle{splncs04}
\bibliography{mapaper-modal-multiverse}

@@

Noetherian set = every infinite sequence everywhere contains repeating terms
https://arxiv.org/pdf/1604.01186.pdf

Cite coquand-persson:groebner

Trees generically absolute as in
https://www.math.uci.edu/~twilson/slides/UCLA-LSC-2013.pdf#page=11

cite https://math.andrej.com/2012/10/03/am-i-a-constructive-mathematician/
